### 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
updated mandc theory paper for section formatting. fig file not working with this template

%%---------------------------------------------------------------------------%%
\section{Conclusion}
These relationships will serve as guidelines for selecting an
appropriate parallel algorithm strategy and provide a basis for
performance models for future parallel implementations of the adjoint
Neumann-Ulam method.

>>>>

%%---------------------------------------------------------------------------%%
\section{Conclusions}
These relationships will serve as guidelines for selecting an
appropriate parallel algorithm strategy and provide a basis for
performance models for future parallel implementations of the adjoint
Neumann-Ulam method.

### 2012-11-30 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updated mandc theory paper for section formatting. fig file not working with this template

%        File: latexdoc.tex
%     Created: Friday March 4 12:24:15 2011
% Last Change: Friday March 4 12:24:20 2011
%
\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[mathcal]{euscript}
\usepackage{tabularx}
\usepackage{cite}
\usepackage{c++}
\usepackage{tmadd,tmath}
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery
  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today}
\title{A Spectral Analysis of the Domain Decomposed Adjoint
  Neumann-Ulam Method}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\begin{abstract}

  The domain decomposed behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the adjoint random walks and the fraction of histories
  leaking from a domain in the decomposition are made with respect to
  the Eigenvalues and other properties of the linear operator. In
  addition, an estimate of the effective optical thickness of a domain
  in the decomposition is presented based on the spectral analysis and
  diffusion theory. Using the leakage fraction estimate, Wigner
  rational approximation and the mean chord approximation are applied
  to estimate the leakage fraction. The one-speed, two-dimensional
  neutron diffusion equation is used as a model problem to test the
  spectral theory for symmetric operators. In general, the spectral
  theory and derived approximations show good agreement with measured
  computational results.

\end{abstract}

%%---------------------------------------------------------------------------%%
\section{Introduction}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method
adjoint to that described by Neumann and Ulam.

@@

For reactor physics Monte Carlo simulations, domain decomposition has
been identified as a key principle in moving forward in high
performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because nearest-neighbor communication
sequences have poor algorithmic strong scaling
\cite{gropp_high-performance_2001}.

For problems where the linear operator is symmetric, a host of
analytic theories exist based on the Eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these theories are
adapted to the domain decomposed adjoint Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation. In this paper
we describe the adjoint Neumann-Ulam method followed by a presentation
and finite difference discretization of the model problem. Using the
linear system generated by this discretization, we use a spectral
theory to generate analytic relations for the Eigenvalues of the
operator based on system parameters. Using the Eigenvalue spectra, we
then build relationships to characterize the transport of random walk
histories in a decomposed domain and compare these analytic results to
numerical experiments conducted with the model problem.

%%---------------------------------------------------------------------------%%
\section{The Adjoint Neumann-Ulam Method}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. Choosing the adjoint method Monte Carlo method to
invert the linear operator, we begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator.  With this
statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
\textit{iteration matrix}.  Using these definitions, we can derive an
estimator from the adjoint method that will also give the solution
vector, $\ve{x}$. We can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of
$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
Eigenvalues and therefore has the same spectral radius. We expand this
summation to again yield a series of transitions that can be
approximated by a Monte Carlo random walk sequence, forming the
Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). The adjoint estimator can be related to
the solution by defining the following inner product equivalence
\cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
require two constraints to close the system. We use
Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source and
the adjoint solution using a delta function source, we recover the
original solution we are looking for. In terms of particle transport, a
delta function source means that this adjoint method is equivalent to
a traditional forward method where tallies are made only in the state
in which the random walk currently resides.

@@

%%---------------------------------------------------------------------------%%
\section{Model Problem}
For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

@@

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the Eigenvalues of the iteration matrix. In
addition, the Eigenvalues of the linear operator from which the
iteration matrix was derived allow for further analysis of convergence
properties. We will compute the Eigenvalues of these matrices by
assuming Eigenfunctions of the form \cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
Eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these Eigenfunctions should
give the Eigenvalues of the matrix as they lie on the unit circle in
the complex plane.

\subsection{Diffusion Operator Spectrum}
For the model problem, we first compute the Eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator to the
Eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

@@

\subsection{Domain Leakage Approximations}
In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\Lambda$ was
dependent on the total cross section of the system via the Wigner
rational approximation. This approximation was developed by Wigner to
compute the neutron leakage probability from the fuel in a reactor as
a simple interpolation between the small and large fuel lump cases,
approximately characterizing the leakage behavior of many common
geometries \cite{duderstadt_nuclear_1976}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  This is a sample LaTeX input file for your contribution to 
%  the MC2013 conference. Modified by R.C. Martineau at INL from A. 
%  Sood at LANL, from J. Wagner ORNL who obtained the original class 
%  file by Jim Warsa, LANL, 16 July 2002}
%
%  Please use it as a template for your full paper 
%    Accompanying/related file(s) include: 
%       1. Document class/format file: mc2013.cls
%       2. Sample Postscript Figure:   figure.eps
%       3. A PDF file showing the desired appearance: template.pdf 
%    Direct questions about these files to: richard.martinea@inl.gov
%
%    Notes: 
%      (1) You can use the "dvips" utility to convert .dvi 
%          files to PostScript.  Then, use either Acrobat 
%          Distiller or "ps2pdf" to convert to PDF format. 
%      (2) Different versions of LaTeX have been observed to 
%          shift the page down, causing improper margins.
%          If this occurs, adjust the "topmargin" value in the
%          mc2013.cls file to achieve the proper margins. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{mc2013}
%
%  various packages that you may wish to activate for usage 
\usepackage{graphicx}
\usepackage{tabls}
\usepackage{afterpage}
\usepackage{cites}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[mathcal]{euscript}
\usepackage{tmadd,tmath}

%\usepackage{epsf}
%
%
% Insert authors' names and short version of title in lines below
%
\newcommand{\authorHead}      % Author's names here
   {Stuart R. Slattery and Paul P.H. Wilson and Thomas M. Evans}  
\newcommand{\shortTitle}      % Short title here
   {Domain-Decomposed Neumann-Ulam Method Analysis}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%   BEGIN DOCUMENT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%
%      Headers and Footers
\afterpage{%
\fancyhf{}%
\fancyhead[CE]{              
{\scriptsize \authorHead}}                                                
\fancyhead[CO]{               
{\scriptsize \shortTitle}}                  
%\lfoot{\scriptsize{
%International Conference on Mathematics and Computational Methods
%Applied to Nuclear Science \& Engineering (M\&C 2013), 
%\\ Sun Valley, Idaho, USA, May 5-9, 2013.}}%
\rfoot{\thepage/\totalpages{}}%

\pagestyle{fancy}
%\setlength{\topmargin}{-20pt}
}

\normalsize

\setlength{\baselineskip}{16.8pt}
\vspace{-3pt}

% 
% TITLE
%

\begin{center}
\textbf{\large \\% 
  A Spectral Analysis of the Domain Decomposed
  Adjoint Neumann-Ulam Monte Carlo Method for Linear Operator
  Equations}
% 
% FIRST AUTHORS 
%
\\
\setlength{\baselineskip}{14pt}
\textbf{Stuart R. Slattery and Paul P.H. Wilson}\footnote{Please send
  correspondence to sslattery@wisc.edu} \\
Engineering Physics Department \\
University of Wisconsin - Madison  \\
1500 Engineering Dr., Madison, WI 53706 \\
sslattery@wisc.edu; wilsonp@engr.wisc.edu \\

% 
% SECOND AUTHORS (if not needed delete from here) 
%
\vspace{12pt}
\textbf{Thomas M. Evans} \\
Radiation Transport Group  \\
Oak Ridge National Laboratory \\
1 Bethel Valley Road, Oak Ridge, TN 37830 \\
evanstm@ornl.gov \\ 
%
% SECOND AUTHORS (to here)
%

\end{center}

%
% SET RAGGED RIGHT MARGIN
%
\raggedright

\section*{ABSTRACT} 
\begin{quote}
\begin{small}
The domain decomposed behavior of the adjoint Neumann-Ulam method for
solving linear operator equations is analyzed using the spectral
properties of the linear operator. Relationships for the average
length of the adjoint random walks and the fraction of histories
leaking from a domain in the decomposition are made with respect to
the Eigenvalues and other properties of the linear operator. In
addition, an estimate of the effective optical thickness of a domain
in the decomposition is presented based on the spectral analysis and
diffusion theory. Using the leakage fraction estimate, Wigner rational
approximation and the mean chord approximation are applied to estimate
the leakage fraction. The one-speed, two-dimensional neutron diffusion
equation is used as a model problem to test the spectral theory for
symmetric operators. In general, the spectral theory and derived
approximations show good agreement with measured computational
results.

\emph{Key Words}: monte carlo domain decomposition adjoint
\end{small} 
\end{quote}

\setlength{\baselineskip}{14pt}
\normalsize

%%---------------------------------------------------------------------------%%
\Section{INTRODUCTION}
\label{sec:intro}

An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method
adjoint to that described by Neumann and Ulam.

@@

To enable higher fidelity reactor physics Monte Carlo simulations,
domain decomposition has been identified as a key principle in moving
forward in high performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because nearest-neighbor communication
sequences have poor algorithmic strong scaling
\cite{gropp_high-performance_2001}.

When solving problems where the linear operator is symmetric, a host
of analytic theories exist based on the Eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these theories are
adapted to the domain decomposed adjoint Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation. In this paper
we describe the adjoint Neumann-Ulam method followed by a presentation
and finite difference discretization of the model problem. Using the
linear system generated by this discretization, we use a spectral
theory to generate analytic relations for the Eigenvalues of the
operator based on system parameters. Using the Eigenvalue spectra, we
then build relationships to characterize the transport of random walk
histories in a decomposed domain and compare these analytic results to
numerical experiments conducted with the model problem.

%%---------------------------------------------------------------------------%%
\Section{THE ADJOINT NEUMANN-ULAM METHOD} 
\label{sec:adjoint_nu}

We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. Choosing the adjoint method Monte Carlo method to
invert the linear operator, we begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator.  With this
statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
\textit{iteration matrix}.  Using these definitions, we can derive an
estimator from the adjoint method that will also give the solution
vector, $\ve{x}$. We can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of
$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
Eigenvalues and therefore has the same spectral radius. We expand this
summation to again yield a series of transitions that can be
approximated by a Monte Carlo random walk sequence, forming the
Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). The adjoint estimator can be related to
the solution by defining the following inner product equivalence
\cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
require two constraints to close the system. We use
Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source and
the adjoint solution using a delta function source, we recover the
original solution we are looking for. In terms of particle transport, a
delta function source means that this adjoint method is equivalent to
a traditional forward method where tallies are made only in the state
in which the random walk currently resides.

@@

%%---------------------------------------------------------------------------%%
\Section{MODEL PROBLEM}
\label{sec:model_problem}

For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
%% \vspace{16pt}
%% \begin{figure}[ht!]
%% \begin{spacing}{1.0}
%%   \begin{center}
%%     \scalebox{1.25}{\input{stencil.pdftex_t}}
%%   \end{center}
%%   \caption{\textbf{Nine-point Laplacian stencil.}}
%%   \label{fig:stencil}
%% \end{spacing}
%% \end{figure}
%% \vspace{16pt}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

@@

%%---------------------------------------------------------------------------%%
\Section{SPECTRAL ANALYSIS}
\label{sec:spectral_analysis}

The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the Eigenvalues of the iteration matrix. In
addition, the Eigenvalues of the linear operator from which the
iteration matrix was derived allow for further analysis of convergence
properties. We will compute the Eigenvalues of these matrices by
assuming Eigenfunctions of the form \cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
Eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these Eigenfunctions should
give the Eigenvalues of the matrix as they lie on the unit circle in
the complex plane.

\subsection{Diffusion Operator Spectrum}
\label{subsec:diffusion_spectrum}

For the model problem, we first compute the Eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator to the
Eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

@@

\subsection{Domain Leakage Approximations}
\label{subsec:domain_leak_approx}

In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\Lambda$ was
dependent on the total cross section of the system via the Wigner
rational approximation. This approximation was developed by Wigner to
compute the neutron leakage probability from the fuel in a reactor as
a simple interpolation between the small and large fuel lump cases,
approximately characterizing the leakage behavior of many common
geometries \cite{duderstadt_nuclear_1976}.
